---
title: "Exploratory Data Analyses"
format: revealjs
editor: visual
author: "Gilles Guillot, PhD"
---

# Illegal wild trade ad data

## 

```{r,echo=FALSE}
library(tidyverse)
iwt = readxl::read_excel("./IWT_test_data.xlsx")
# shorter variable names
iwt = iwt %>% rename(website=`website/name`,
                             type = `sale-item/type`,
                             species = `sale-item/species`,
                             taxa = `sale-item/taxa` , 
                             sold_in = `sold-in`,
                             CITES = `sale-item/CITES`)
# type in species names
iwt = iwt %>% mutate(species = fct_collapse(species, 
                                            Bear = c('bear','Bear')))
iwt = iwt %>% mutate(species = fct_collapse(species, 
                                            `Radiated Tortoise` = c('Radiated tortoise','Radiated Tortoise')))

```

* Products mostly advertised on Facebook
* Products mostly sold in  Thailand
* Top species include Elephant, Tiger & Bear

```{r, echo = FALSE}
iwt %>% 
  ggplot(aes(y=species,fill=website)) + 
  geom_bar() + 
  facet_wrap(~ sold_in)

```

## 

Marked seasonal trend in Thailand

```{r}
iwt %>% group_by(sold_in, month) %>% 
  summarise(n=n()) %>% 
  mutate(month = factor(month,levels = c('August','September', 'October'))) %>% 
 ggplot(aes(x=month , y=n, col=sold_in)) +
  geom_point(size=5,alpha=.7)
```



## Other potentially informative data


* Weight, size, volume, # animals involved
* Website sampling/census strategy
* Price
* Clue that transaction took place

## Conclusions: 

* Policy implications: target effort on Thailand in October

* Data visualization plan: 
   + Elicit research question
   + Translate research question into statistical relations
   + Play with ggplot() *ad libitum*


# Brazil deforestation data



```{r,echo =FALSE}
# data loading and formatting
library(pacman)
p_load("geobr")
library(geobr)
library(sf)

# forest_dat municipalities
forest_dat = readxl::read_excel("./Brazilian_Municipalities_test_data.xlsx")

# downloading Brazil geographical data
muni <- geobr::read_municipality(  code_muni = "all",
                            year=2020, 
                            showProgress = FALSE)
my_list_muni = unique(forest_dat$municipality_code)
my_muni = muni %>% filter(code_muni %in% my_list_muni)


# map Brazil + GI_TOC test data
my_muni = my_muni %>% rename(municipality_code = code_muni)


# merging geobr data and GI_TOC test data
dat = full_join(x = forest_dat , y = my_muni, by = "municipality_code") %>% 
  st_as_sf()

# Computing new variables
dat = dat %>% mutate(municipality_area = st_area(geom)) %>% 
  mutate(municipality_area_km2 = as.vector(municipality_area/1e+6)) %>% 
  mutate(deforestation_rate = 100*deforestation_km2 / municipality_area_km2) %>% 
  mutate(pop_density = count_population / municipality_area_km2) %>% 
  mutate(rural_conflicts_per_1000inhab = 1000 * count_rural_conflicts / count_population) %>% 
  mutate(rural_conflicts_per_km2 = count_rural_conflicts / municipality_area_km2)

```

## Study area: Amazon basin
```{r}
ggplot() + 
  geom_sf(data = muni, color=NA, fill = 'forestgreen') + 
  geom_sf(data=my_muni,fill ="dodgerblue3")
```


## Choosing the right denominator

* Municipality  surfaces are highly variable
* Deforestation surface has to be scaled to municipality surface:
$$ \mbox{Deforestation rate} = \frac{ \mbox{Deforested surface}}{ \mbox{Municipality surface}} \times 100$$

* What is the right denominator to scale  rural conflicts?


## Spatial variation of deforestation and  rural conflict rates 

```{r}
# map defor rate and rural conflict rate
dat %>% select(deforestation_rate,rural_conflicts_per_1000inhab,geom) %>% 
  pivot_longer(cols=c("deforestation_rate","rural_conflicts_per_1000inhab")) %>% 
ggplot() + 
  geom_sf(aes(fill = value)) +
  facet_wrap(~name) + 
  scale_fill_distiller(palette = "Spectral")
```

## Deforestation vs. rural conflict rates 

```{r}
dat %>% ggplot(aes(x=deforestation_rate, y = rural_conflicts_per_1000inhab)) +  
  geom_point(size=4,alpha=.3, col="red4") + 
  xlab("Deforestation rate (%)") + 
  ylab("Rural conflicts per 1000 inhabitant") + 
  scale_x_log10() + 
   scale_y_log10() 

```

## Auxiliary analyses

* Deforestation is "easy" to detect and monitor from remote sensing data

* Is there an association between deforestation and (other) criminal activities, e.g.
   + deforestation ~ count_homicide
    + deforestation ~ count_seizures_CITES
    

## Selected publications

*  Population dynamics of species-rich ecosystems: the mixture of matrix population models approach. In: [Methods in Ecology and Evolution 4.4 (2013), pp. 316–326.](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210x.12019)
*  Les éléphants en danger. In: [Brèves de maths: Mathématiques pour la planète terre. Nouveau Monde éditions](http://www.breves-de-maths.fr/les-elephants-en-danger/), 2014.
* Statistical methods in spatial genetics. In: [Molecular Ecology 18 (2009), pp. 4734–4756](https://onlinelibrary.wiley.com/doi/10.1111/j.1365-294X.2009.04410.x). 

## Selected publications (cont')

* Accurate continuous geographic assignment from low-to high-density SNP data. In:
[Bioinformatics 32.7 (2016), pp. 1106–1108](https://academic.oup.com/bioinformatics/article/32/7/1106/1744090).
*  Enhancing and comparing methods for the detection of fishing activity from Vessel Monitoring System data. In: [arXiv preprint arXiv:1708.09663 (2017)](https://arxiv.org/abs/1708.09663).



# Thank you for your attention!

All data  material available from my github repository[github.com/gilles-guillot/GI-TOC](https://github.com/gilles-guillot/GI-TOC)
